---
header-includes:
   - \providecommand{\titletxt}{Tutorial - Spatial Analysis in R with Open Geodata}
   - \usepackage[skins]{tcolorbox}
   - \usepackage{float}
   - \usepackage{multicol}
   - \usepackage{graphicx}
   - \usepackage{lastpage}
   - \usepackage{fancyhdr}
   - \pagestyle{fancy}
   - \fancyhf{}
   - \lhead{Use of R in Official Statistics - uRos2018}
   - \rhead{6th international conference, The Hague, September 2018}
   - \usepackage[numbered]{bookmark}
   - \usepackage{hyperref}
   - \hypersetup{
     pdftitle={Tutorial - Spatial Analysis in R with Open Geodata - uRos2018},
     pdfauthor={Egge-Jan Pollé and Willy Tadema}
     }
   - \hypersetup{linkcolor = {blue}, urlcolor = {blue}}
   - \renewcommand{\footrulewidth}{0.4pt}
   - \fancyfoot{}
   - \fancyfoot[l]{\titletxt}
   - \fancyfoot[r]{Page \thepage\ of \pageref*{LastPage}}
   - \usepackage{fvextra}
   - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
   - \usepackage{xcolor}
   - \definecolor{grey}{rgb}{0.8,0.8,0.8}
   - \usepackage{enumitem}
   - \renewcommand{\figurename}{Figure}
output: 
  pdf_document:
    latex_engine: lualatex
mainfont: Calibri
urlcolor: blue
---


\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE,echo = TRUE)
```

\huge Tutorial - Spatial Analysis in R with Open Geodata

\normalsize

**Egge-Jan Pollé** - Tensing GIS Consultancy B.V.[^1]  
**Willy Tadema** - Provincie Groningen[^2]

Version 0.1.0 - July 31, 2018

\begin{multicols}{2}

\subsection{Introduction}


This is the manual for a tutorial session to be delivered on Wednesday 12 September 2018 at the 6\textsuperscript{th} International \href{https://www.aanmelder.nl/uros2018}{Conference on the Use of R in Official Statistics (uRos2018)} at the Dutch Office for National Statistics (\href{https://www.cbs.nl/en-gb}{CBS}) in the Hague.

\subsection{Prerequisites}

\begin{itemize}
  \item Attendees should bring their own laptop with (64-bit) \href{https://www.r-project.org/}{R}, and preferably also \href{https://www.rstudio.com/}{RStudio}, installed.
  \item Shortly before the training session we will publish a list of additional packages (e.g. \textbf{sf}, \textbf{mapview} and \textbf{tmap}) to be installed - please keep an eye on our \href{https://github.com/TWIAV/Spatial_Analysis_in_R_with_Open_Geodata}{GitHub repository}.
  \item No specific prior experience (with R) is required for this introductory session, though some basic experience in one or more of the following fields is certainly helpful: Data Science, Programming/Software Development and/or GIS.
\end{itemize}

\end{multicols}

[^1]: https://www.linkedin.com/in/ejhpolle/
[^2]: https://www.linkedin.com/in/willytadema/

##Course material available on GitHub

The URL of this repository: [https://github.com/TWIAV/Spatial_Analysis_in_R_with_Open_Geodata](https://github.com/TWIAV/Spatial_Analysis_in_R_with_Open_Geodata)

\newpage
\newpage
\pdfbookmark{\contentsname}{toc}
\hypersetup{linkcolor=black}
\setcounter{secnumdepth}{4}
\setcounter{tocdepth}{4}
\tableofcontents
\hypersetup{linkcolor=blue}
\newpage

# Managing Geospatial Vector Data in R {-}  

# The package `sf` (Simple Features for R) \label{simple_features}

To manage spatial data in R you will use the library `sf`, a relativle new addition to the R universe. The package was [released on CRAN in January 2017](https://www.r-consortium.org/blog/2017/01/03/simple-features-now-on-cran).

This package provides support for simple features, which is a standardized way to encode spatial vector data.

`sf` links directly to three important geospatial libraries, to unlock their power for use in R:

* GDAL for reading and writing data
* GEOS for geometrical operations
* Proj.4 for projection conversions and datum transformations.

**The package `sf` on CRAN:**  
[https://cran.r-project.org/web/packages/sf/index.html](https://cran.r-project.org/web/packages/sf/index.html)


If the package is not yet installed, you can install it with the following command:

```{r install_sf, eval = FALSE}
install.packages("sf")
```

To be able to manage your spatial data you will first have to load the package `sf`:

```{r library_sf}
library(sf)
```

\tcbset{coltitle=black, colbacktitle=white, colframe=gray!75!black,colback=white,nobeforeafter}
\begin{tcolorbox}[enlarge by=5mm, hyphenationfix, title=The real geospatial powers behind sf]
\begin{itemize}
\item GDAL: the \textbf{Geospatial Data Abstraction Library} is a translator library for raster and vector geospatial data formats. Website: \url{http://www.gdal.org/}
\item GEOS: the \textbf{Geometry Engine, Open Source} contains the complete functionality of the OpenGIS Simple Features for SQL spatial predicate functions and spatial operators. Website: \url{https://trac.osgeo.org/geos}
\item Proj.4: \textbf{PROJ} is a generic coordinate transformation software, that transforms coordinates from one coordinate reference system (CRS) to another. This includes cartographic projections as well as geodetic transformations. Website: \url{http://proj4.org/}
\end{itemize}
\end{tcolorbox}

# A first exercise

```{r}
ap <- "Lelystad Airport"
cd <- "LEY"
class(cd)
lat <- 52.460278
lon <- 5.527222
class(lon)
airport <- data.frame(ap, cd, lat, lon, stringsAsFactors = FALSE)
NL_Airports <- read.csv("http://www.twiav.nl/files/NL_Airports.csv", stringsAsFactors = FALSE)

print(NL_Airports)

```
```{r, eval = FALSE}
NL_Airports <- rbind(NL_Airports, airport) # This will give an error
```
```{r}
names(airport) <- names(NL_Airports)

NL_Airports <- rbind(NL_Airports, airport)
print(NL_Airports)
class(NL_Airports)

library(sf)

NL_Airports <- st_as_sf(NL_Airports, coords = c("longitude","latitude"), crs = 4326)

class(NL_Airports)

plot(st_geometry(NL_Airports), pch = 17)

```

\newpage

# Interactive Viewing of Spatial Data in R

Until now you have seen some static maps plotted o the **Plots** pane. But as a real Data Scientist you also want to be able to explore your data on an interactive map. Until a few years ago this would have meant switching back and forth between a desktop GIS and R. But in recent years some new packages have been developed to enable interactive map viewing.

In this chapter we will look at some of these.

## The package `mapview`

**The package `mapview` on CRAN:**  
[https://cran.r-project.org/web/packages/mapview/index.html](https://cran.r-project.org/web/packages/mapview/index.html)

If the package is not yet installed, you can install it with the following command:

```{r install_mv, eval = FALSE}
install.packages("mapview")
```

To be able to view your spatial data interactively you will first have to load the package `mapview`:

```{r library_mv}
library(mapview)
```


\newpage

# Accessing Geospatial Vector Data over the Internet {-}

# Downloadable shapefiles

\begin{multicols}{2}

In this chapter we will learn how to download, to unzip and to load shapefiles \textbf{using R}. There will be no need to use your web browser, your file explorer or a zip utility - the whole process can be completed using just a few lines of R code.

The shapefile format is a well-known and still rather popular geospatial vector data format for Geographic Information System (GIS) software. It spatially describes vector features - points, lines, and polygons - with attribute data attached. The shapfile is a 'classic' GIS file format in the sense that it stores geometry and attribute data in separate files (in this chapter we will discover that a single shapefile in reality consits of multiple files) as opposed to more modern spatial database or file formats, where geometry and attributes are stored together in a single table or file.

\includegraphics{Images/cbs.png}

The shapefile format has been developed by \href{https://www.esri.com/}{Esri} and over the years has become a \textit{de facto} standard for data interoperability among Esri and other GIS software products.

It is not uncommon for public organizations - including national statistical institutes - to distribute geographic information in this shapefile format. So, you will find shapefiles on the \href{https://www.cbs.nl/nl-nl/dossier/nederland-regionaal/geografische-data}{page with geographic data of the Dutch Office for National Statistics (CBS)} and also on the \href{http://data.statistik.gv.at/web/catalog.jsp}{STATISTIK AUSTRIA open.data Portal}.

\includegraphics{Images/logo_statistik.png}

\end{multicols}

## Download and unzip the shapefile

```{r eval=FALSE}
# Set the working directory
setwd("C:/uRos2018")
# Store the URL to the file to download in a variable
URL2zip <- "http://data.statistik.gv.at/data/OGDEXT_GEM_1_STATISTIK_AUSTRIA_20180101.zip"
print(URL2zip)
# Extract the filename from this URL
zip_file <- file.path(basename(URL2zip))
print(zip_file)
# Create a subfolder in your working directory to store the data
dir.create("./Data", showWarnings = FALSE)
# store the path to the current working directory in a variable
workdir <- getwd()
# switch to the data folder you just created
setwd("./Data")
# Download the file
download.file(URL2zip, destfile = zip_file, mode = "wb")
# unzip the file
unzip(zip_file)
# after unzipping you can delete (i.e. unlink) the file
unlink(zip_file)
# Return to the original working directory
setwd(workdir)
# remove variables you do not longer need
rm(URL2zip,zip_file,workdir)
```

```{r echo=FALSE}
setwd("C:/uRos2018")
URL2zip <- "http://data.statistik.gv.at/data/OGDEXT_GEM_1_STATISTIK_AUSTRIA_20180101.zip"
zip_file <- file.path(basename(URL2zip))
dir.create("C:/uRos2018/Data", showWarnings = FALSE)
setwd("C:/uRos2018/Data")
download.file(URL2zip, destfile = zip_file, mode = "wb")
unzip(zip_file)
unlink(zip_file)
setwd("C:/uRos2018")
```

## Load the shapefile


```{r eval=FALSE}
library(sf)
library(tmap)
AUSTRIA_GEM_20180101 <- st_read("./Data/STATISTIK_AUSTRIA_GEM_20180101.shp")
```

```{r echo=FALSE}
library(sf)
library(tmap)
AUSTRIA_GEM_20180101 <- st_read("C:/uRos2018/Data/STATISTIK_AUSTRIA_GEM_20180101.shp")
```

```{r eval=FALSE}
qtm(AUSTRIA_GEM_20180101)
```

```{r echo=FALSE}
qtm(AUSTRIA_GEM_20180101)
```



\newpage

# OGC Web Feature Service (WFS) \label{chapWFS}

## Introduction

\begin{multicols}{2}

In this chapter we will learn how \textbf{to use R as a client to access data using a Web Feature Service (WFS)}. WFS is a Data Access Standard which is defined and maintained by the Open Geospatial Consortium (OGC). The WFS Interface Standard defines a set of interfaces for accessing geographic information over the Internet. It offers the means to retrieve geographic features and their properties through a highly configurable interface and in a manner independent of the underlying data stores they publish.

The standard is used - both in the public and private sector and in the academic world - to publish vector geospatial datasets in a way that makes it easy for receiving organisations to conduct analysis on the data supplied.

In a short chapter like this it will not be possible to cover all ins and outs of the WFS Interface Standard. The main goal here is to get you up and running and to whetten your appetite for more. It might all look a bit technical and intimidating at first, but as soon as you have the syntax of your request right, you will be able to retrieve valuable spatial data in the blink of an eye.

\end{multicols}

## WFS Outputformat: GML vs. GeoJSON \label{gmlGEOJSON}

By default, a WFS returns data in Geography Markup Language (GML) which is written as eXtensible Markup Language (XML). However, many WFS services also offer the option to request the ouput in GeoJSON, a geospatial data interchange format based on JavaScript Object Notation (JSON).

By it's very nature, GML data is difficult to process, because - as with most XML based grammars - there are two parts to the grammar: the schema that describes the document and the instance document that contains the actual data.

On the contrary, the GeoJSON standard clearly defines several types of JSON objects and the manner in which they are combined to represent data about geographic features, their properties, and their spatial extents.

In general **GDAL**, the translator library behind the `sf` functions `st_read()` and `st_write` (see chapter \ref{simple_features}), gives better results with GeoJSON as opposed to GML.

So, in the exercises in this manual, when accessing a WFS service o retrieve data, we will always add the parameter `outputFormat=application/json`.

## Access a WFS service: `request=GetCapabilities` \label{GCnl}

In general, organisations publishing data using WFS, will provide you with a URL to their WFS server which also contains a `GetCapabilities` reguest. Once you know the **capabilities** of the service (i.e. once you know what is on offer) you can start building your own request to the server.

In the exercises below we will use data on the municipal division in the Netherlands. These data are offered by the host of this conference, the Dutch statistical office, through [Publieke Dienstverlening op de Kaart](https://www.pdok.nl/) (PDOK) and the [Nationaal Georegister](http://www.nationaalgeoregister.nl/) (NGR).

Some background information on the dataset used, can be found here: [Dataset: CBS Gebiedsindelingen](https://www.pdok.nl/introductie?articleid=1951759).

The actual URL with the `GetCapabilities` reguest is:  
[https://geodata.nationaalgeoregister.nl/cbsgebiedsindelingen/wfs?request=GetCapabilities](https://geodata.nationaalgeoregister.nl/cbsgebiedsindelingen/wfs?request=GetCapabilities)

When you click this link the server response will be shown in your browser in XML format.

We are not going to study this XML response line-by-line. Not now. But please leave the web page open in your browser for later reference.

## Retrieve data from a WFS service: `request=GetFeature` \label{GetFeaturePar}

### `request=GetFeature` - an example from the Netherlands \label{GFnl}

A WFS server responding to a *GetFeature* request returns a collection of geographic feature instances filtered according to a criteria set by the requesting client.

We will start with a simple request to download a full feature collection without any constraints to filter the content by. The GetFeature request queries the server with a set of parameters, which are concatenated to the URL with an *ampersand* (&).

In the script below we have chosen to store the different parts of our request in separate variables, only to concatenate them at the end to get the full URL. We do so for readability reasons and to allow for easy modification of our request at a later stage if necessary.

A WFS service can offer one or more feature collections, see the `<FeatureTypeList>` section in the XML response to the `GetCapalities` request. The service we are accessing here offers quite some feature collections, i.e. multiple regional divisions for the years 1995 up to the current year.

What we are intersted in here, is the municipal division for the year 2017. After some browsing through the XML response, we have found a `<FeatureType>` with the `<Name>` **cbsgebiedsindelingen:cbs_gemeente_2017_gegeneraliseerd**. And that's the value we are giving to the `typename` parameter. (This `typenames` parameter determines the collection of feature instances to return.)

Aslso, do not forget to add a parameter to ask for output in GeoJSON format (as discussed in paragraph \ref{gmlGEOJSON}).

This is the script to populate the full request:

```{r}
wfs_baseurl <- "https://geodata.nationaalgeoregister.nl/cbsgebiedsindelingen/wfs?"
wfs_service <- "service=WFS"
wfs_version <- "version=2.0.0"
wfs_request <- "request=GetFeature"
wfs_typename <- "typename=cbsgebiedsindelingen:cbs_gemeente_2017_gegeneraliseerd"
wfs_outputformat <- "outputFormat=application/json"

wfs_fullurl <- paste0(wfs_baseurl, wfs_service,"&",wfs_version,"&",wfs_request,"&",wfs_typename,"&",wfs_outputformat)
print(wfs_fullurl)

```

The variable `wfs_fullurl` will now contain this [link](https://geodata.nationaalgeoregister.nl/cbsgebiedsindelingen/wfs?service=WFS&version=2.0.0&request=GetFeature&typename=cbsgebiedsindelingen:cbs_gemeente_2017_gegeneraliseerd&outputFormat=application/json). (When you click this link the server response will be shown in your browser in GeoJSON format.)

Now we want to feed this response directly into R, like this:

```{r}
library(sf)
library(tmap)

NL_Municipalities2017 <- st_read(wfs_fullurl)
```

\newpage
Have a look at the result:

```{r}

qtm(NL_Municipalities2017)

head(NL_Municipalities2017)

```
\newpage

### `request=GetFeature` - an example from Finland


In this paragraph we will retrieve data from [Tilastokeskus (Statistics Finland)](http://www.stat.fi/). In the english section of their website we found a nice dataset: [Population by municipality-based units](http://www.stat.fi/org/avoindata/paikkatietoaineistot/vaesto_tilastointialueittain_en.html).


\begin{multicols}{2}
And from this map it is especially the layer \textbf{Population 2017 by municipalities 2018 (kunta\_vaki2017)} we would like to investigate.

We will follow the same steps as in the Dutch example above. For an explanation of the steps taken, please refer to paragraph \ref{GCnl} and \ref{GFnl}.

\includegraphics{Images/tklogo_fi.png}

\end{multicols}

The URL with `GetCapabilities` request is:  
[http://geo.stat.fi/geoserver/vaestoalue/wfs?request=GetCapabilities](http://geo.stat.fi/geoserver/vaestoalue/wfs?request=GetCapabilities)

In this XML response - in the `<FeatureTypeList>` section - we have found a `<FeatureType>` with the `<Name>` **vaestoalue:kunta_vaki2017**. That's the value we are giving to the `typename` parameter.

The script to populate te full URL is similar to the one before - the only variables we have modified are `wfs_baseurl` and `wfs_typename`:

```{r}
# FI: Example with Finnish data
wfs_baseurl <- "http://geo.stat.fi/geoserver/vaestoalue/wfs?"
wfs_service <- "service=WFS"
wfs_version <- "version=2.0.0"
wfs_request <- "request=GetFeature"
wfs_typename <- "typename=vaestoalue:kunta_vaki2017"
wfs_outputformat <- "outputFormat=application/json"

wfs_fullurl <- paste0(wfs_baseurl, wfs_service,"&",wfs_version,"&",wfs_request,"&",wfs_typename,"&",wfs_outputformat)
print(wfs_fullurl)

```

The variable `wfs_fullurl` will now contain this [link](http://geo.stat.fi/geoserver/vaestoalue/wfs?service=WFS&version=2.0.0&request=GetFeature&typename=vaestoalue:kunta_vaki2017&outputFormat=application/json). (When you click this link the server response will be shown in your browser in GeoJSON format.)

Now we want to feed this response directly into R, like this:

```{r}
library(sf)
library(tmap)

FI_Municipalities2018_Pop2017 <- st_read(wfs_fullurl)

```
\newpage

Have a look at the result:

```{r}
qtm(FI_Municipalities2018_Pop2017)
```

Optionally, you can have a look at the attribute data in the **Data Viewer**: `View(FI_Municipalities2018_Pop2017)`

\begin{multicols}{2}

With some basic knowledge of the Finnish language you should now be able to calculate that Finland had more ore less 5.5 million inhabitants in 2017 - of which more or less 50 percent are males, and the other half females:

\tcbset{coltitle=black, colbacktitle=white, colframe=gray!75!black,colback=white,nobeforeafter}
\begin{tcolorbox}[enlarge by=5mm, hyphenationfix, title=Finnish for Data Scientists: some useful words]
\begin{itemize}
\item \textbf{Miehet} = Men
\item \textbf{Naiset} = Women
\item \textbf{Nimi} = Name
\item \textbf{Vaesto} = Population
\end{itemize}
\end{tcolorbox}

\end{multicols}

```{r}
sum(FI_Municipalities2018_Pop2017$vaesto)
sum(FI_Municipalities2018_Pop2017$miehet)
sum(FI_Municipalities2018_Pop2017$naiset)
```

\newpage

## Get a description of a dataset: `request=DescribeFeatureType` \label{parDescFeatType}

In the previous exercises we immediately executed `GetFeature` requests to WFS services - mainly because the trainer told us to do so - without actually knowing anything about these datasets.

To get a description of the dataset, you can execute a `DescribeFeatureType` request first. This returns a description - in XML format - of the structure, including properties, of the feature type specified in the request.

For the Dutch municipalities dataset this request would look like this:
[https://geodata.nationaalgeoregister.nl/cbsgebiedsindelingen/wfs?service=WFS&version=2.0.0&request=DescribeFeatureType&typename=cbsgebiedsindelingen:cbs_gemeente_2017_gegeneraliseerd](https://geodata.nationaalgeoregister.nl/cbsgebiedsindelingen/wfs?service=WFS&version=2.0.0&request=DescribeFeatureType&typename=cbsgebiedsindelingen:cbs_gemeente_2017_gegeneraliseerd)

And for the Finnish dataset like this:
[http://geo.stat.fi/geoserver/vaestoalue/wfs?service=WFS&version=2.0.0&request=DescribeFeatureType&typename=vaestoalue:kunta_vaki2017](http://geo.stat.fi/geoserver/vaestoalue/wfs?service=WFS&version=2.0.0&request=DescribeFeatureType&typename=vaestoalue:kunta_vaki2017)

## Some additional parameters to the `GetFeature` request

Additional parameters can be added to a `GetFeature` request to further filter or convert the response from the WFS.

To include additional parameters to a request, simply add an *ampersand* (&) at the end of the URL, then add the name of the parameter, an equal sign (=) and the value to assign to the parameter.

Below we will discuss a few of the parameters available.

### Limit the number of records returned: the `count` parameter

To have a look at the structure of a dataset - before retrieving the full dataset - you can choose to limit the number of records returned with the `count` parameter.

The `GetFeature` request below limits the number of Finnish municipalities returned to only 5:

```{r}
# FI: Example with Finnish data
wfs_baseurl <- "http://geo.stat.fi/geoserver/vaestoalue/wfs?"
wfs_service <- "service=WFS"
wfs_version <- "version=2.0.0"
wfs_request <- "request=GetFeature"
wfs_typename <- "typename=vaestoalue:kunta_vaki2017"
wfs_count <- "count=5"
wfs_outputformat <- "outputFormat=application/json"

wfs_fullurl_w_count <- paste0(wfs_baseurl, wfs_service,"&",wfs_version,"&",wfs_request,"&",wfs_typename,"&",wfs_count,"&",wfs_outputformat)
print(wfs_fullurl_w_count)

```
The variable `wfs_fullurl_w_count` will now contain this [link](http://geo.stat.fi/geoserver/vaestoalue/wfs?service=WFS&version=2.0.0&request=GetFeature&typename=vaestoalue:kunta_vaki2017&count=5&outputFormat=application/json). (When you click this link the server response will be shown in your browser in GeoJSON format.)

\newpage

### Limit the number of rows returned: the `PropetyName` parameter

To restrict a `GetFeature` request by attribute rather than feature, use the `PropertyName` parameter. You can specify a single attribute, or multiple attributes separated by commas.

The `GetFeature` request below only returns the geometry and the columns *nimi* (Name) and *vaesto* (total population) for the Finnish municipalities:

```{r}
# FI: Example with Finnish data
wfs_baseurl <- "http://geo.stat.fi/geoserver/vaestoalue/wfs?"
wfs_service <- "service=WFS"
wfs_version <- "version=2.0.0"
wfs_request <- "request=GetFeature"
wfs_typename <- "typename=vaestoalue:kunta_vaki2017"
wfs_propertyname <- "propertyname=geom,nimi,vaesto"
wfs_outputformat <- "outputFormat=application/json"

wfs_fullurl_w_properties <- paste0(wfs_baseurl, wfs_service,"&",wfs_version,"&",wfs_request,"&",wfs_typename,"&",wfs_propertyname,"&",wfs_outputformat)
print(wfs_fullurl_w_properties)

```
The variable `wfs_fullurl_w_properties` will now contain this [link](http://geo.stat.fi/geoserver/vaestoalue/wfs?service=WFS&version=2.0.0&request=GetFeature&typename=vaestoalue:kunta_vaki2017&propertyname=geom,nimi,vaesto&outputFormat=application/json). (When you click this link the server response will be shown in your browser in GeoJSON format.)

\newpage

## Further reading

The OGC offers an official tutorial module - [OGC E-learning](http://cite.opengeospatial.org/pub/cite/files/edu/index.html) - covering it's activities and the several standards it maintains. The [specific chapter on WFS](http://cite.opengeospatial.org/pub/cite/files/edu/wfs/text/basic-index.html) can be found [here](http://cite.opengeospatial.org/pub/cite/files/edu/wfs/text/basic-index.html).

\newpage

# Esri ArcGIS Online

\begin{multicols}{2}

\href{https://developers.arcgis.com/rest/}{ArcGIS REST API}

\href{https://livingatlas.arcgis.com/en/}{ArcGIS Living Atlas of the World}

Some background information on the Living Atlas of the World can be found in this \href{http://www.esri.com/esri-news/arcuser/summer-2015/welcome-to-the-living-atlas-of-the-world}{article}, where Esri also invites YOU to \textit{"Become a user and a contibutor"}.



\includegraphics{Images/living-atlas-logo-size-4.png}

\end{multicols}

\newpage

## Exploring the data in ArcGIS Online

Once you have found a nice (and open) dataset in the Living Atlas, you can click to proceed to an overview page with metadata. For the dataset we have chosen - the **Feature Layer** *Les communes en France 2018* - this overview page can be found via this [link](https://www.arcgis.com/home/item.html?id=a3325fff0a3343e0becb44bf27317970#overview) (see Figure \ref{FR_Com18_AGOL_Metadata}). 

In [Map Viewer](https://www.arcgis.com/home/webmap/viewer.html?useExisting=1&layers=a3325fff0a3343e0becb44bf27317970)

\newpage

\begin{figure}[H]
\centering
\caption{On the overview page metadata about \textbf{Feature Layer} \textit{Les communes en France 2018} is presented}
\vspace{5pt}
\includegraphics{Images/Feature_Layer_Communes_France_2018_Metadata.png}
\label{FR_Com18_AGOL_Metadata}
\end{figure}
\begin{figure}[H]
\centering
\caption{Here the \textbf{Feature Layer} \textit{Les communes en France 2018} is shown in the \textbf{ArcGIS Online Map Viewer}}
\vspace{5pt}
\includegraphics{Images/Feature_Layer_Communes_France_2018_Map_Viewer.png}
\label{FR_Com18_AGOL_MapViewer}
\end{figure}

\newpage

## Query the Feature Service

[https://services.arcgis.com/d3voDfTFbHOCRwVR/arcgis/rest/services/COMMUNE/FeatureServer/0](https://services.arcgis.com/d3voDfTFbHOCRwVR/arcgis/rest/services/COMMUNE/FeatureServer/0)

[https://services.arcgis.com/d3voDfTFbHOCRwVR/arcgis/rest/services/COMMUNE/FeatureServer/0/query](https://services.arcgis.com/d3voDfTFbHOCRwVR/arcgis/rest/services/COMMUNE/FeatureServer/0/query)

\newpage

\begin{figure}[H]
\centering
\caption{On this page the \textbf{Feature Layer} \textit{COMMUNE} is presented in the \textbf{ArcGIS REST Services Directory}}
\vspace{5pt}
\includegraphics{Images/ArcGIS_REST_Services_Directory_Commune_FeatureServer.png}
\label{FR_Com18_AG_REST_Services_Dir}
\end{figure}
\begin{figure}[H]
\centering
\caption{On this page you can enter parameters to query the \textbf{Feature Layer} \textit{COMMUNE}}
\vspace{5pt}
\includegraphics{Images/ArcGIS_REST_Services_Directory_Commune_FeatureServer_query.png}
\label{FR_Com18_AG_REST_Services_Dir_Query}
\end{figure}


\newpage

# References {-}

* **Pebesma, Edzer** (January 3, 2017). *Simple Features Now on CRAN*. Blog. Retreived from: [https://www.r-consortium.org/blog/2017/01/03/simple-features-now-on-cran](https://www.r-consortium.org/blog/2017/01/03/simple-features-now-on-cran)


\newpage

# Appendix A: List of abbreviations used {-}

Abbreviation | Meaning
-------------|----------
CRAN | Comprehensive R Archive Network
GIS | Geographic Information System
GML | Geography Markup Language
JSON | JavaScript Object Notation
OGC | Open Geospatial Consortium
WFS | Web Feature Service
XML | eXtensible Markup Language

\newpage

# Appendix B: Additional exercises {-}

## OGC Web Feature Service (WFS) {-}

For background information on the steps taken in the exercise below, please refer to chapter \ref{chapWFS}

### `request=GetFeature` - another example from the Netherlands {-}

In this additional exercise we will access a WFS service offering information about *bevolkingskernen* (urban human settlements) in the Netherlands in 2011. More information on this dataset can be found here: [Dataset: CBS Bevolkingskernen](https://www.pdok.nl/introductie?articleid=1951745).

The URL with `GetCapabilities` request is:  
[https://geodata.nationaalgeoregister.nl/bevolkingskernen2011/wfs?request=GetCapabilities](https://geodata.nationaalgeoregister.nl/bevolkingskernen2011/wfs?request=GetCapabilities)

In the `<FeatureTypeList>` section of this XML response we can see that this service only offers one `<FeatureType>`, i.e. one dataset. This dataset has the `<Name>` **bevolkingskernen2011:cbsbevolkingskernen2011**. That's the value we are giving to the `typename` parameter in our `GetFeature` request.


The URL with `DescribeFeatureType` request is:  
[https://geodata.nationaalgeoregister.nl/bevolkingskernen2011/wfs?service=WFS&version=2.0.0&request=DescribeFeatureType&typename=bevolkingskernen2011:cbsbevolkingskernen2011](https://geodata.nationaalgeoregister.nl/bevolkingskernen2011/wfs?service=WFS&version=2.0.0&request=DescribeFeatureType&typename=bevolkingskernen2011:cbsbevolkingskernen2011)

This dataset does contain quite some attribute columns!

This is the script to populate the full `GetFeature` request:

```{r}
# NL: Urban Human Settlement Analysis
wfs_baseurl <- "https://geodata.nationaalgeoregister.nl/bevolkingskernen2011/wfs?"
wfs_service <- "service=WFS"
wfs_version <- "version=2.0.0"
wfs_request <- "request=GetFeature"
wfs_typename <- "typename=bevolkingskernen2011:cbsbevolkingskernen2011"
wfs_outputformat <- "outputFormat=application/json"

wfs_fullurl <- paste0(wfs_baseurl, wfs_service,"&",wfs_version,"&",wfs_request,"&",wfs_typename,"&",wfs_outputformat)
print(wfs_fullurl)

```
The variable `wfs_fullurl` will now contain this [link](https://geodata.nationaalgeoregister.nl/bevolkingskernen2011/wfs?service=WFS&version=2.0.0&request=GetFeature&typename=bevolkingskernen2011:cbsbevolkingskernen2011&outputFormat=application/json). (When you click this link the server response will be shown in your browser in GeoJSON format.)

Now we want to feed this response directly into R, like this:

```{r}
library(sf)

NL_Urban_Human_Settlements2011 <- st_read(wfs_fullurl)
```



\newpage

```{r}
plot(st_geometry(NL_Urban_Human_Settlements2011), col = "orange", border = "red")
```

